Factors affecting reliablity of state-space age-structured assessment models

### 2025-02-13 Project_0/manuscript/manuscript.Rmd
saving work on project_0 manuscript

---
title: Factors affecting reliablity of state-space age-structured assessment models
author: 
  - Timothy J. Miller^1,2^
  - Greg Britten^3^
  - Elizabeth N. Brooks^2^
  - Gavin Fay^4^
  - Alex Hansell^2^
  - Christopher M. Legault^2^
  - Brandon Muffley^5^
  - Brian C. Stock^6^
  - John Wiedenmann^7^
date: "`r format(Sys.time(), '%d %B, %Y')`"
fontsize: 12pt
output:
  pdf_document:
#    citation_package: natbib
    fig_caption: yes
    includes:
      in_header:
      - options.sty
      - journalshortcuts.tex
    keep_tex: yes
    number_sections: yes
  html_document:
    df_print: paged
#  word_document: default
header-includes:
#- \usepackage{makecell}
csl: canadian-journal-of-fisheries-and-aquatic-sciences.csl
bibliography: manuscript.bib
#biblio-style: cjfas2.bst
---

@@

Random effects can be included in state-space assessment models on many processes and in many ways, and guidance is needed on statistical reliability and model selection criteria. We simulated 72 operating models with varying fishing history, observation error uncertainty, and process error magnitude, correlation, and source (recruitment, survival, fishery selectivity, catchability, and natural mortality). We fit estimating models with different assumptions on the process error source, whether median natural mortality was estimated, and whether a stock-recruit relationship was estimated.

Models that assumed the correct process error source had high convergence and low bias. Bias was also low under most process error assumptions when there was contrast in fishing pressure. Stock-recruitment parameters were only reliably estimated in ideal situations. Marginal AIC most accurately distinguished process errors on recruitment, survival, and selectivity, as well as larger magnitude process errors of other types. Retrospective patterns were generally weak except for recruitment when observation error was high, even with the correct process error assumptions. When models did exhibit some retrospective pattern, estimating median natural mortality often removed it.

@@

Much is known about the reliability of state-space models that are linear or Gaussian [@aeberhardetal18], but applications in fisheries management are nonlinear and typically include multiple types of observations with varying distributional assumptions. Furthermore, there is a wide range of potential random effects structures and model parameters that can be treated as random effects in assessment models. We know relatively little about the factors affecting statistical reliability of such models or the ability of information criteria to distinguish among such alternative structures.

@lietal24 investigated some aspects of inferences for operating models with multiple sources of process error, but there are differences for this paper.

@@

Here we conduct a simulation study with operating models (OMs) varying by degree of observation error uncertainty, source and variability of process error, and fishing history. The simulations from these OMs are fitted with estimation models (EMs) that make alternative assumptions for sources of process error, whether a stock-recruit model was estimated, and whether M is estimated. We evaluate whether AIC can correctly determine the correct source of process error and stock effects on recruitment. We also evaluate the degree of bias in the outputs of the assessment model that are important for management.

@@

We used the Woods Hole Assessment Model (WHAM) to configure operating and EMs in our simulation study [@millerstock20;@stockmiller21]. WHAM is an R package freely available as a github repository. For this study we used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3). This package has also been used to configure operating and EMs for closed loop simulations evaluating index-based assessment methods [@legaultetal23] and is used for management of haddock, butterfish, American plaice, bluefish, Atlantic cod, black sea bass, and yellowtail flounder in the Northeast US.

@@

For R and R+S OMs, convergence generally declines for most EMs when the median natural mortality rate is estimated and/or the Beverton-Holt stock recruit relationship is estimated even when the process error assumptions of the EMs and OMs match (Figure \ref{hessian_SE_convergence}, A). When there was high observation error and constant fishing pressure (change from $2.5F_{\text{MSY}}$ to $F_{\text{MSY}}$), convergence of all EM process error configurations other than R EMs was poor for R OMs ($\sigma_{2+} = 0$) regardless of whether median natural mortality and stock-recruit relationships were estimated. Convergence of R EMs was high for all R and R+S OMs except when there was high observation error and constant fishing pressure and median natural mortality and stock-recruit relationships were estimated.  R+S EMs fit to R OMs exhibited poor convergence regardless of whether natural mortality or a stock-recruit relationship was estimated. R+S EMs fit to R+S OMs had highest convergence rates when there was contrast in fishing pressure and low observation error. Convergence rates were high for all EMs when fit to data from R+S OMs with lower observation uncertainty except those where median natural mortality and/or stock-recruit relationships were estimated.

@@

R+S EMs, in particular, converged poorly when fit to data generated from R+Sel OMs with lower selectivity process error variability or higher observation uncertainty (Figure \ref{hessian_SE_convergence}, C). R+Sel EMs generally converged better than other EMs for R+Sel OMs with higher process error variability, lower observation uncertainty, and contrast in fishing pressure  regardless of whether median natural mortality or a stock recruit relationship was estimated.

@@

Marginal AIC accurately determines the correct process error assumptions in EMs when data are generated from R and R+S OMs, regardless of whether median natural mortality or a stock recruit relationship is estimated (Figure \ref{pe_aic}, A). Adding estimation of median natural mortality or a stock recruit relationship separately has a negligible effect on the accuracy of determining the correct process error assumption. When both are estimated, there is a noticeable reduction in accuracy when OMs have a constant fishing history, observation uncertainty is low and larger variability in recruitment process errors.

@@

Marginal AIC most accurately determined the correct source of process error and correlation structure for R+Sel OMs with low observation uncertainty (Figures \ref{pe_aic}, C). When there was low variability in selectivity process errors and high observation uncertainty, R+q or R+S EMs were more likely to have the best AIC. Whether median natural morality or stock recruit relationships were estimated appeared to have little effect on the performance of AIC.

Marginal AIC most accurately determined the correct source of process error and correlation structure for R+q OMs with high variability in catchability process errors (Figures \ref{pe_aic},D). The worst accuracy occurred for OMs with low variability in catchability process errors and high observation uncertainty. However, in these OMs, the marginal AIC accurately determined the correct source of process error (but not correlation structure) except when EMs estimated both median natural morality and the stock recruit relationships and OMs assumed a constant fishing pressure.

@@

With R and R+S process error assumptions, probability of AIC accuracy for the B-H stock-recruit relationship as a function of SSB variability were greatest for OMs where there was greater process variability in survival, lower process variability in recruitment and
contrast in fishing pressure (Figure \ref{sr_aic}, A). However, the largest variation in SSB (Figure \ref{sr_aic}, rug on x-axis) occurred in OMs with larger recruitment variability ($\sigma_R = 1.5$; Figure \ref{sr_aic}, A, right column group). The weakest overall accuracy of AIC for the Beverton-Holt stock-recruit function was when there was constant fishing pressure (Figure \ref{sr_aic}, A, second and fourth rows). The largest effect of simultaneously estimating median natural mortality was when there was OMs had high observation uncertainty and constant fishing pressure (Figure \ref{sr_aic}, A, second row).

For R+M OMs, probability of AIC accuracy for the Beverton-Holt stock-recruit relationship was greatest for OMs where there was contrast in fishing pressure (Figure \ref{sr_aic}, B). The largest variation in SSB occurred in OMs with larger natural mortality variability and temporal AR1 correlation of the annual random effects ($\sigma_M = 1.5$, $\rho_M = 0.9$; Figure \ref{sr_aic}, B, right-most column) where AIC accuracy was also high even when there was constant fishing pressure. Interestingly, there seemed to be little difference in AIC accuracy whether there was lower or higher observation uncertainty or whether the EM simultaneously estimated median natural mortality.

@@

Like the R+Sel OMs, the greatest accuracy for AIC in selecting the Beverton-Holt stock-recruit relationship occurred for R+q OMs where there was contrast in fishing pressure over time which is also where there was the greatest variation in SSB (Figure \ref{sr_aic}, D, first and third rows). There were some differences in effects of SSB variation on AIC accuracy when OMs had high observation uncertainty and constant fishing pressure. Across all of the OMs in Figure \ref{sr_aic}, the only estimated decline in AIC accuracy with SSB variation occurred when R+q OMs had high observation uncertainty and constant fishing pressure and greater variation in catchability process errors that also had AR1 autocorrelation.

@@

For R OMs ($\sigma_{2+} = 0$), there was no indication of bias (95\% confidence intervals included 0) in terminal year SSB for any of the estimating models regardless of process error assumptions, except when no SR assumption was made, recruitment variability was low, and there was contrast in fishing mortality and high observation uncertainty (Figure \ref{SSB_rel_error}, A). However, errors in terminal SSB estimates were highly variable when median natural mortality was estimated and there was constant fishing pressure and high observation uncertainty (Figure \ref{SSB_rel_error}, A, second row).

For R+S OMs, the EMs with matching process error assumption generally produced unbaised estimation of terminal SSB except when median natural mortality was estimated and there was high observation uncertainty. In R+S OMs with low observation uncertainty, EMs with incorrect process error assumptions typically provided biased estimation of terminal year SSB. Estimating the Beverton-Holt stock-recruit relationship had little discernible effect on bias of terminal year SSB estimation whereas estimating median M tended to produce more variability in errors in terminal SSB estimation similar to R OMs.

For R+ M OMs with low variability in natural mortality process errors, low observation uncertainty and contrast in fishing motality over time all EMs produced low variability in SSB estimation error that indicated unbiasedness (Figure \ref{SSB_rel_error}, B, third row). However, larger variability in natural mortality process errors increased bias of EMs without the correct process error type. Estimating median natural mortality increased variability of SSB estimation error particularly for OMs with high observation uncertainty and constant fishing pressure over time and increased bias in SSB estimation for many R+M OMs. Like R and R+S OMs, estimating a stock-recruit relationship had little discernible effect on SSB bias.

For R+Sel OMs, there was no evidence of bias for any EMs when variability in selectivity process errors was low, observation uncertainty low and contrast in fishing mortality (Figure \ref{SSB_rel_error}, C). The largest bias for all EMs that estimated median natural mortality when the OMs had high observation uncertainty, constant fishing pressure and greater variability in selectivity process errors ($\sigma_{\text{Sel}} = 0.5$) or low selectivity process errors ($\sigma_{\text{Sel}} = 0.1$) and low observation uncertainty. However, there was no evidence of bias of matching R+Sel EMs when observation uncertainty was low and variation in selectivity process errors was larger, whether median natural mortality was estimated or not (Figure \ref{SSB_rel_error}, C, third row). We only observed an effect of estimating the Beverton-Holt Stock recruit relationship for R+Sel OMs that had high observation uncertainty and contrast in fishing pressure where estimating the relationship produced less biased SSB estimation for many EMs (Figure \ref{SSB_rel_error}, C, top row).

All EMs fit to data from R+q OMs with low observation uncertainty and  contrast in fishing pressure exhibited little evidence of bias in terminal SSB estimation except for R+M EMs when there was no AR1 correlation in catchability process errors (Figure \ref{SSB_rel_error}, D). Many EMs also performed well in R+q OMs with low observation uncertainty, but no contrast in fishing pressure. For R+q OMs with high observation uncertainty and contrast in fishing pressure, EMs that estimated the Beverton-Holt stock recruit function exhibited less SSB bias than those that did not. Estimating median natural mortality in the EMs only resulted in much more variable SSB estimation errors when there was no contrast in fishing pressure (Figure \ref{SSB_rel_error}, D, first and third rows).

@@

Bias for stock-recruit parameters was very strong and variability in relative errors was greatest for most EMs fit to R+Sel OMs with constant fishing pressure (Figure \ref{SR_rel_error}, C). Less bias in parameter estimation occurred for OMs with a change in fishing pressure over time and of those OMs and the most accurate estimation occurred for EMs with process error types that matched the OMs and the OMs had low observation uncertainty, contrast in fishing mortality and more variable and uncorrelated selectivity process errors. There was little effect of estimating natural mortality on relative errors for stock recruit parameters.

Like R+Sel OMs, relative errors in stock-recruit parameters were less variable for most EM process error types when OMs had contrast in fishing pressure and those OMs that also had lower observation uncertainty generally had the least variability (Figure \ref{SR_rel_error}, D). The worst accuracy of stock-recruit parameter estimation regardless of EM type occured when R+q OMs had low observation uncertainty and constant fishing pressure (Figure \ref{SR_rel_error}, D, fourth row).

@@

Median natural mortality rate is estimated accurately by all EM process error types for all R OMs except those where observation uncertainty is high and fishing pressure is constant where variability in relative errors is high (Figure \ref{M_rel_error}, A, $\sigma_{2+} = 0$).  For R+S OMs estimation of median natural mortality rate is most accurate when observation uncertainty is low and there is contrast in fishing pressure and the EM process error type is correct. There is little effect of estimating stock-recruit relationships on the bias for median natural mortality.

For R+M OMs, median natural mortality was estimated most accurately, regardless of EM process error type, when OMs had a change in fishing pressure and low observation uncertainty (Figure \ref{M_rel_error}, B). However, those R+M OMs that also had greatest variability in AR1 correlated natural mortality process errors only had unbiased estimation when the EM process error type was correct. Like R, and R+S OMs, there was little effect of estimating stock-recruit relationships on the bias for median natural mortality.

All EM process error types produced accurate estimation of median natural mortality rate for R+Sel OMs that had contrast in fishing pressure, low observation uncertainty, and low selectivity process error variability (Figure \ref{M_rel_error}, C). When selectivity process error variability increased the incorrect EM process errors produce more biased estimation of median natural mortality rate. The least accurate estimation occurred for all EM process error types when observation uncertainty was high and there was constant fishing pressure. Again, like other OM process error types, there was little effect of estimating stock-recruit relationships on the bias for median natural mortality.

Like R+Sel OMs, all EM process error types produced accurate estimation of median natural mortality rate when fit to R+q OMs with contrast in fishing pressure, low observation uncertainty and low catchability process error variability (Figure \ref{M_rel_error}, D). Most EM process error types produced biased estimation of median natural  mortality when R+q OMs had high observaiton uncertainty and constant fishing pressure. Again, there was no discernible effect of estimating stock-recruit relationships on bias.

@@

Mohn's $\rho$ for SSB was small in absolute value for all R and R+S OMs, regardless of EM process error types, and whether median natural mortality rate or stock-recruit relationships were estimated (Figure \ref{mohns_rho_ssb}, A). The strongest retrospective patterns occurred in OMs with the largest survival process error variability, high observation uncertainty, and contrast in fishing pressure, but only for EMs without the correct process error type and where median natural mortality rate was assumed known. For R+M, R+Sel, and R+q OMs, Mohn's $\rho$ was also small in absolute value, but median values were all closer to 0 than the largest values in the R and R+S OMs (Figure \ref{mohns_rho_ssb},B-D). For these OMs, there was no noticeable effect of estimation of median natural mortality rate or stock recruit relationship on Mohn's $\rho$ for any EM process error types.

Mohn's $\rho$ for recruitment was small in absolute value for all R OMs with low variability in recruitment process errors, regardless of EM process error type, and whether median natural mortality rate or stock-recruit relationships were estimated (Figure \ref{mohns_rho_R}, A). However, R and R+S OMs with greater recruitment process variability and higher observation uncertainty had median Mohn's $\rho$ for recruitment greater than zero for most EMs even when the the EM process error type was correct. In R+S OMs with lower observation uncertainty, EMs with the correct process error type exhibited better median Mohn's $\rho$ close to 0 than EMs with the incorrect process error type. For R+M, R+Sel, and R+q OMs, results for Mohn's $\rho$ for recruitment are similar to those for SSB, but the range in median values and variation in Mohn's $\rho$ values for a given OM are generally larger for recruitment (Figure \ref{mohns_rho_R}, B-D).

@@

Convergence results can be useful for understanding how lack of convergence in applications to real data might direct the practitioner to which alternative random effects configurations is more appropriate. Therefore, the type of convergence that we might use as a diagnostic is important. It is common during assessment model fitting to check that the maximum absolute gradient component is less than some threshold, but there is no standard. However, checking the Hessian for positive definiteness is sometimes done conditional on the convergence criterion. As such, some models may be excluded that in fact have an invertible hessian. We found the hessian at the optimized log-likelihood can often be invertible when the gradient approach with a sensible criterion (e.g., $10^{-3}$) would indicate lack of convergence.

Another factor affecting the convercence criteria is numerical accuracy. With the process errors modeled as random effects we are optimizing an (Laplace) approximation of the marginal log-likelihood and therefore hessians and gradients are also with regard to the approximation. Furthermore, numerical methods are used to calculate the hessian (Optimhess?), and also to invert it for variance estimation. All of these approximations are affecting the gradients and invertibility of the hessian (although likely worse for some models than others), which, along with our results, suggests at least investigating whether the hessian is positive definite when the calculated gradients are not terribly large.

Often high gradients and/or lack of positive definite hessian result when selectivity parameter estimates are at their bounds (0,1) or random effects variance parameter estimates are essentially 0. Changing the parameter configuration to avoid the bounds or removing process errors with 0 variance will often lead to satisfactory convergence.

Perhaps a greater gradient threshold would be more consistent with hessian-based convergence, but the gradient and hessian-based standard errors should be inspected.

Parameters that are not well estimated can lead the assessment scientist toward a more appropriate and often more parsimonious model configuration. For example, variance parameters that are estimated on log scale as low negative number (variance is close to 0) can happen for random effects estimated with little or no variability. These may sometimes result in a hessian that is not positive definite or due to numerical issues may be invertible, but provide very large standard errors. Generally, we can expect lower probability of convergence when estimating natural mortality or stock-recruit relationships because of the difficulty distinguishing these parameters from others being estimated in assessment model with data typically available.

@@

Among the process error configurations we used in OMs, we found AIC to be accurate for process errors on recruitment and survival (R and R+S). Fitting models to other OMs rarely preferred R+S EMs and R and R+S EMs were nearly always selected for the matching OMs. For other sources of process error, accuracy of AIC was improved to useful levels when there was larger variability in the process errors and/or better data quality.

>>>>

---
title: Factors affecting reliablity of state-space age-structured assessment models
author: 
  - Timothy J. Miller^1,2^
  - Greg Britten^3^
  - Elizabeth N. Brooks^2^
  - Gavin Fay^4^
  - Alex Hansell^2^
  - Christopher M. Legault^2^
  - Chengxue Li^2^
  - Brandon Muffley^5^
  - Brian C. Stock^6^
  - John Wiedenmann^7^
date: "`r format(Sys.time(), '%d %B, %Y')`"
fontsize: 12pt
output:
  pdf_document:
#    citation_package: natbib
    fig_caption: yes
    includes:
      in_header:
      - options.sty
      - journalshortcuts.tex
    keep_tex: yes
    number_sections: yes
  html_document:
    df_print: paged
#  word_document: default
header-includes:
#- \usepackage{makecell}
csl: canadian-journal-of-fisheries-and-aquatic-sciences.csl
bibliography: manuscript.bib
#biblio-style: cjfas2.bst
---

@@

State-space assessment models, which are increasingly being used for management of commercially important fish stocks, can include process errors in many ways, and an evaluation of the statistical reliability of the estimates of management output and best practices of selection among alternative process error configurations are needed. We simulated 72 operating models with varying fishing pressure history, observation error error, and process error magnitude, correlation, and source (recruitment, survival, fishery selectivity, catchability, and natural mortality). We fit estimating models with different assumptions on the process error source, whether median natural mortality was estimated, and whether a stock-recruit relationship was estimated. Estimating models that assumed the correct process error source and median natural mortality rate was known, and did not estimate a stock recruit relationship generally had high probability of convergence and low bias in terminal year SSB estimation. Bias was also low under many incorrect process error assumptions when there was contrast in fishing pressure and low observation error. Stock-recruitment parameters were not reliably estimated in any scenarios, but the least accurate estimates occured under ideal situations of low observation error, low recruitment variability, and temporal contrast in fishing pressure. However, estimation of median natural mortality rate was reliable for many OMs (other than those with survival process errors) with temporal contrast in fishing pressure regardless of the estimated process error type. Marginal AIC most accurately distinguished process errors on recruitment, survival, and selectivity, as well as larger magnitude process errors of other types. Retrospective patterns for spawning stock biomass and fishing mortality were generally weak, but those for recruitment were sizable when observation error was high, even with the correct process error assumptions.

@@

Although the statistical aspects of state-space models and their application have been studied extensively, the work is primarily on Gaussian state-space models. State-space stock assessment models, where there are nonlinear functions of latent parameters and multiple types of observations with varying distributional assumptions, is one of the most complex examples of this analytical approach.

Much is known about the reliability of state-space models that are linear or Gaussian [@aeberhardetal18], but not for the more complex application in stock assessment modeling.

[@lietal24 investigated some aspects of inferences for operating models with multiple sources of process error, but there are differences for this paper.]

There is a wide range of model parameters that can be treated as random effects in assessment models and a variety of potential distributions for the random effects. We know relatively little about the factors affecting statistical reliability of such models or the ability to distinguish among such alternative structures.

@@

Here we conduct a simulation study with operating models (OMs) varying by degree of observation error error, source and variability of process error, and fishing history. The simulations from these OMs are fitted with estimation models (EMs) that make alternative assumptions for sources of process error, whether a stock-recruit model was estimated, and whether a constant, or, in some EMs, median, natural mortality is estimated. We evaluate whether AIC can correctly determine the correct source of process error and a stock-recruit relationship. We also evaluate when retrospective patterns occur and the degree of bias in the outputs of the assessment model that are important for management.

@@

We used the Woods Hole Assessment Model (WHAM) to configure OMs and EMs in our simulation study [@millerstock20;@stockmiller21]. WHAM is an R package freely available as a github repository. For this study we used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3). This package has also been used to configure OMs and EMs for closed loop simulations evaluating index-based assessment methods [@legaultetal23] and is used for management of haddock, butterfish, American plaice, bluefish, Atlantic cod, black sea bass, and yellowtail flounder in the Northeast US.

@@

For R and R+S OMs, convergence generally declines for most EMs when the median natural mortality rate is estimated and/or the Beverton-Holt stock recruit relationship is estimated even when the process error assumptions of the EMs and OMs match (Figure \ref{hessian_SE_convergence}, A). When there was high observation error and constant fishing pressure (change from $2.5F_{\text{MSY}}$ to $F_{\text{MSY}}$), convergence of all EM process error configurations other than R EMs was poor for R OMs ($\sigma_{2+} = 0$) regardless of whether median natural mortality and stock-recruit relationships were estimated. Convergence of R EMs was high for all R and R+S OMs except when there was high observation error and constant fishing pressure and median natural mortality and stock-recruit relationships were estimated.  R+S EMs fit to R OMs exhibited poor convergence regardless of whether natural mortality or a stock-recruit relationship was estimated. R+S EMs fit to R+S OMs had highest convergence rates when there was contrast in fishing pressure and low observation error. Convergence rates were high for all EMs when fit to data from R+S OMs with lower observation error except those where median natural mortality and/or stock-recruit relationships were estimated.

@@

R+S EMs, in particular, converged poorly when fit to data generated from R+Sel OMs with lower selectivity process error variability or higher observation error (Figure \ref{hessian_SE_convergence}, C). R+Sel EMs generally converged better than other EMs for R+Sel OMs with higher process error variability, lower observation error, and contrast in fishing pressure  regardless of whether median natural mortality or a stock recruit relationship was estimated.

@@

Marginal AIC accurately determines the correct process error assumptions in EMs when data are generated from R and R+S OMs, regardless of whether median natural mortality or a stock recruit relationship is estimated (Figure \ref{pe_aic}, A). Adding estimation of median natural mortality or a stock recruit relationship separately has a negligible effect on the accuracy of determining the correct process error assumption. When both are estimated, there is a noticeable reduction in accuracy when OMs have a constant fishing history, observation error is low and larger variability in recruitment process errors.

@@

Marginal AIC most accurately determined the correct source of process error and correlation structure for R+Sel OMs with low observation error (Figures \ref{pe_aic}, C). When there was low variability in selectivity process errors and high observation error, R+q or R+S EMs were more likely to have the best AIC. Whether median natural morality or stock recruit relationships were estimated appeared to have little effect on the performance of AIC.

Marginal AIC most accurately determined the correct source of process error and correlation structure for R+q OMs with high variability in catchability process errors (Figures \ref{pe_aic},D). The worst accuracy occurred for OMs with low variability in catchability process errors and high observation error. However, in these OMs, the marginal AIC accurately determined the correct source of process error (but not correlation structure) except when EMs estimated both median natural morality and the stock recruit relationships and OMs assumed a constant fishing pressure.

@@

With R and R+S process error assumptions, probability of AIC accuracy for the B-H stock-recruit relationship as a function of SSB variability were greatest for OMs where there was greater process variability in survival, lower process variability in recruitment and
contrast in fishing pressure (Figure \ref{sr_aic}, A). However, the largest variation in SSB (Figure \ref{sr_aic}, rug on x-axis) occurred in OMs with larger recruitment variability ($\sigma_R = 1.5$; Figure \ref{sr_aic}, A, right column group). The weakest overall accuracy of AIC for the Beverton-Holt stock-recruit function was when there was constant fishing pressure (Figure \ref{sr_aic}, A, second and fourth rows). The largest effect of simultaneously estimating median natural mortality was when there was OMs had high observation error and constant fishing pressure (Figure \ref{sr_aic}, A, second row).

For R+M OMs, probability of AIC accuracy for the Beverton-Holt stock-recruit relationship was greatest for OMs where there was contrast in fishing pressure (Figure \ref{sr_aic}, B). The largest variation in SSB occurred in OMs with larger natural mortality variability and temporal AR1 correlation of the annual random effects ($\sigma_M = 1.5$, $\rho_M = 0.9$; Figure \ref{sr_aic}, B, right-most column) where AIC accuracy was also high even when there was constant fishing pressure. Interestingly, there seemed to be little difference in AIC accuracy whether there was lower or higher observation error or whether the EM simultaneously estimated median natural mortality.

@@

Like the R+Sel OMs, the greatest accuracy for AIC in selecting the Beverton-Holt stock-recruit relationship occurred for R+q OMs where there was contrast in fishing pressure over time which is also where there was the greatest variation in SSB (Figure \ref{sr_aic}, D, first and third rows). There were some differences in effects of SSB variation on AIC accuracy when OMs had high observation error and constant fishing pressure. Across all of the OMs in Figure \ref{sr_aic}, the only estimated decline in AIC accuracy with SSB variation occurred when R+q OMs had high observation error and constant fishing pressure and greater variation in catchability process errors that also had AR1 autocorrelation.

@@

For R OMs ($\sigma_{2+} = 0$), there was no indication of bias (95\% confidence intervals included 0) in terminal year SSB for any of the estimating models regardless of process error assumptions, except when no SR assumption was made, recruitment variability was low, and there was contrast in fishing mortality and high observation error (Figure \ref{SSB_rel_error}, A). However, errors in terminal SSB estimates were highly variable when median natural mortality was estimated and there was constant fishing pressure and high observation error (Figure \ref{SSB_rel_error}, A, second row).

For R+S OMs, the EMs with matching process error assumption generally produced unbaised estimation of terminal SSB except when median natural mortality was estimated and there was high observation error. In R+S OMs with low observation error, EMs with incorrect process error assumptions typically provided biased estimation of terminal year SSB. Estimating the Beverton-Holt stock-recruit relationship had little discernible effect on bias of terminal year SSB estimation whereas estimating median M tended to produce more variability in errors in terminal SSB estimation similar to R OMs.

For R+ M OMs with low variability in natural mortality process errors, low observation error and contrast in fishing motality over time all EMs produced low variability in SSB estimation error that indicated unbiasedness (Figure \ref{SSB_rel_error}, B, third row). However, larger variability in natural mortality process errors increased bias of EMs without the correct process error type. Estimating median natural mortality increased variability of SSB estimation error particularly for OMs with high observation error and constant fishing pressure over time and increased bias in SSB estimation for many R+M OMs. Like R and R+S OMs, estimating a stock-recruit relationship had little discernible effect on SSB bias.

For R+Sel OMs, there was no evidence of bias for any EMs when variability in selectivity process error and observation error was low, and with contrast in fishing mortality (Figure \ref{SSB_rel_error}, C). The largest bias occurred for any EMs that estimated median natural mortality when the OMs had high observation error, constant fishing pressure, and greater variability in selectivity process errors ($\sigma_{\text{Sel}} = 0.5$) or low selectivity process errors ($\sigma_{\text{Sel}} = 0.1$) and low observation error. However, there was no evidence of bias of matching R+Sel EMs when observation error was low and variation in selectivity process errors was larger, whether median natural mortality was estimated or not (Figure \ref{SSB_rel_error}, C, third row). We only observed an effect of estimating the Beverton-Holt Stock recruit relationship for R+Sel OMs that had high observation error and contrast in fishing pressure where estimating the relationship produced less biased SSB estimation for many EMs (Figure \ref{SSB_rel_error}, C, top row).

All EMs fit to data from R+q OMs with low observation error and  contrast in fishing pressure exhibited little evidence of bias in terminal SSB estimation except for R+M EMs when there was no AR1 correlation in catchability process errors (Figure \ref{SSB_rel_error}, D). Many EMs also performed well in R+q OMs with low observation error, but no contrast in fishing pressure. For R+q OMs with high observation error and contrast in fishing pressure, EMs that estimated the Beverton-Holt stock recruit function exhibited less SSB bias than those that did not. Estimating median natural mortality in the EMs only resulted in much more variable SSB estimation errors when there was no contrast in fishing pressure (Figure \ref{SSB_rel_error}, D, first and third rows).

@@

Bias for stock-recruit parameters was very strong and variability in relative errors was greatest for most EMs fit to R+Sel OMs with constant fishing pressure (Figure \ref{SR_rel_error}, C). Less bias in parameter estimation occurred for OMs with a change in fishing pressure over time and of those OMs and the most accurate estimation occurred for EMs with process error types that matched the OMs and the OMs had low observation error, contrast in fishing mortality and more variable and uncorrelated selectivity process errors. There was little effect of estimating natural mortality on relative errors for stock recruit parameters.

Like R+Sel OMs, relative errors in stock-recruit parameters were less variable for most EM process error types when OMs had contrast in fishing pressure and those OMs that also had lower observation error generally had the least variability (Figure \ref{SR_rel_error}, D). The worst accuracy of stock-recruit parameter estimation regardless of EM type occured when R+q OMs had low observation error and constant fishing pressure (Figure \ref{SR_rel_error}, D, fourth row).

@@

Median natural mortality rate is estimated accurately by all EM process error types for all R OMs except those where observation error is high and fishing pressure is constant where variability in relative errors is high (Figure \ref{M_rel_error}, A, $\sigma_{2+} = 0$).  For R+S OMs estimation of median natural mortality rate is most accurate when observation error is low and there is contrast in fishing pressure and the EM process error type is correct. There is little effect of estimating stock-recruit relationships on the bias for median natural mortality.

For R+M OMs, median natural mortality was estimated most accurately, regardless of EM process error type, when OMs had a change in fishing pressure and low observation error (Figure \ref{M_rel_error}, B). However, those R+M OMs that also had greatest variability in AR1 correlated natural mortality process errors only had unbiased estimation when the EM process error type was correct. Like R, and R+S OMs, there was little effect of estimating stock-recruit relationships on the bias for median natural mortality.

All EM process error types produced accurate estimation of median natural mortality rate for R+Sel OMs that had contrast in fishing pressure, low observation error, and low selectivity process error variability (Figure \ref{M_rel_error}, C). When selectivity process error variability increased the incorrect EM process errors produce more biased estimation of median natural mortality rate. The least accurate estimation occurred for all EM process error types when observation error was high and there was constant fishing pressure. Again, like other OM process error types, there was little effect of estimating stock-recruit relationships on the bias for median natural mortality.

Like R+Sel OMs, all EM process error types produced accurate estimation of median natural mortality rate when fit to R+q OMs with contrast in fishing pressure, low observation error and low catchability process error variability (Figure \ref{M_rel_error}, D). Most EM process error types produced biased estimation of median natural  mortality when R+q OMs had high observaiton error and constant fishing pressure. Again, there was no discernible effect of estimating stock-recruit relationships on bias.

@@

Mohn's $\rho$ for SSB was small in absolute value for all R and R+S OMs, regardless of EM process error types, and whether median natural mortality rate or stock-recruit relationships were estimated (Figure \ref{mohns_rho_ssb}, A). The strongest retrospective patterns occurred in OMs with the largest survival process error variability, high observation error, and contrast in fishing pressure, but only for EMs without the correct process error type and where median natural mortality rate was assumed known. For R+M, R+Sel, and R+q OMs, Mohn's $\rho$ was also small in absolute value, but median values were all closer to 0 than the largest values in the R and R+S OMs (Figure \ref{mohns_rho_ssb},B-D). For these OMs, there was no noticeable effect of estimation of median natural mortality rate or stock recruit relationship on Mohn's $\rho$ for any EM process error types.

Mohn's $\rho$ for recruitment was small in absolute value for all R OMs with low variability in recruitment process errors, regardless of EM process error type, and whether median natural mortality rate or stock-recruit relationships were estimated (Figure \ref{mohns_rho_R}, A). However, R and R+S OMs with greater recruitment process variability and higher observation error had median Mohn's $\rho$ for recruitment greater than zero for most EMs even when the the EM process error type was correct. In R+S OMs with lower observation error, EMs with the correct process error type exhibited better median Mohn's $\rho$ close to 0 than EMs with the incorrect process error type. For R+M, R+Sel, and R+q OMs, results for Mohn's $\rho$ for recruitment are similar to those for SSB, but the range in median values and variation in Mohn's $\rho$ values for a given OM are generally larger for recruitment (Figure \ref{mohns_rho_R}, B-D).

@@

Convergence results can be useful for understanding how lack of convergence in applications to real data might direct the practitioner to which alternative random effects configurations is more appropriate. Therefore, the type of convergence that we might use as a diagnostic is important. It is common during assessment model fitting to check that the maximum absolute gradient component is less than some threshold, but there is no standard. However, if the Hessian were checked for positive definiteness conditional on a satisfactory gradient, some models may be excluded that in fact have an invertible hessian. We found the hessian at the optimized log-likelihood can often be invertible when the gradient approach with a sensible criterion (e.g., $10^{-3}$) would indicate lack of convergence.

Another factor affecting the convergence criteria is numerical accuracy. With the process errors modeled as random effects we are optimizing an (Laplace) approximation of the marginal log-likelihood and therefore hessians and gradients are also with regard to the approximation. Furthermore, numerical methods are used to calculate the hessian (Optimhess?), and also to invert it for variance estimation. All of these approximations are affecting the gradients and invertibility of the hessian (although likely worse for some models than others), which, along with our results, suggests at least investigating whether the hessian is positive definite when the calculated gradients are not terribly large. Often high gradients and/or lack of positive definite hessian result when parameter estimates are at their bounds [@carvalhoetal21], and this also applies to variance or autocorrelation parameters for random effects with state-space assessment models.

Even when hessian-based variance can be accomplished, parameters that are poorly estimated will have extremely large variances. This further inspection can lead to a more appropriate and often more parsimonious model configuration where the problematic parameters are not estimated. For example, process error variance parameters that are estimated close to 0 indicates that the random effects are estimated to have little or no variability and removing these process errors is warranted. Generally, our results suggest we can expect lower probability of convergence of state-space assessment models when estimating natural mortality or stock-recruit relationships because of the difficulty distinguishing these parameters from others being estimated in assessment model with data typically available.

@@

Among the process error configurations we used in OMs, we found AIC to be accurate for process errors on recruitment and survival (R and R+S). Fitting models to other OMs rarely preferred R+S EMs and R and R+S EMs were nearly always selected for the matching OMs. For other sources of process error, accuracy of AIC was improved to useful levels when there was larger variability in the process errors and/or lower observation error.

We found AIC to performs weakly in determining stock recruit relationships unless there is large contrast in SSB and low variation in recruitment process errors.  Although we did not compare models with alternative stock-recruit relationships (e.g., Ricker and Beverton-Holt), we do not expect AIC to perform any better distinguishing between relationships. Our finding that AIC tended to choose simpler recruitment models in most cases contrasts with the noted bias in AIC for more complex models [@shibata76;@katz81;@kassraftery95], but whereas those findings apply to more common fitting of models to raw data, state-space models account for observation error and estimate process errors in latent variables.
